The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-08-11 manuscript/rebuttal.Rmd

Dear Editor,

Many thanks for your consideration of our manuscript.
Below is our point-by-point response to reviewer comments.

### Reviewer 1

> In this manuscript, the authors present a framework that computational researchers should follow to
> maximize the reproducibility of their computational research. The five pillars that they identified
> are 1) code version control and sharing, 2) compute environment control,  3) literate programming,
> 4) documentation and 5) FAIR data sharing. They give a fair amount of detail regarding each pillar.
> This paper contains no original research and is more of an opinion piece. With that in mind, if BiB
> is interested in this topic, I recommend publishing. The authors make very reasonable suggestions.
> My only request for edits is to rearrange the section structure of the manuscript to be in line with
> the numbering of the pillars. Right now it is quite confusing.

Response: We thank the reviewer for their positive feedback.
With regards to the ordering of the five pillars, we have amended Figure 1 so that the pillars now
appear in the same order as they are detailed in sections following.

@@

### Reviewer 2

> Comments to the Author: The article offers a comprehensive perspective on reproducibility in
> computational research, delivering clear and logically backed recommendations. It is indeed a very
> interesting topic that needs our attention. Nevertheless, there are areas that could benefit from
> further development and clarity:

@@

Response: We appreciate reviewer 2's positive feedback and thoughtful suggestions.

> 1. Inclusion of Case Studies or Examples: To strengthen your argument, consider incorporating case
> studies or practical examples that demonstrate the effectiveness of your recommendations.
> Highlighting specific instances where these best practices have enhanced reproducibility or solved
> particular issues could substantiate your points more effectively.

@@

Consider case study of enrichment analysis.

> 2. Expanding on Current Shortcomings: The article mentions that current practices are not meeting the
> goal of reproducibility. An expanded discussion on the ramifications of these shortcomings on
> scientific progress could accentuate the urgency and significance of addressing this issue.

@@

* Challenges

* Conclusion

* Key points

> 5. Engaging Conclusion: Reworking the conclusion to be more impactful could be beneficial. Instead of
> merely summarizing the recommendations, it would be useful to indicate the potential trajectory of
> computational research if these recommendations are adopted, as well as highlighting the potential
> risks if they aren't.

### 2023-08-11 manuscript/5pillars.Rmd
potti case study

Much has been written on how irreproducibility has reached crisis levels in psychology and life
sciences [@Baker2015; @Begley2012-di].
Indeed, in bioinformatics, the situation is grim with a 2009 systematic evaluation showing only
2 of 18 articles could be reproduced (11%) @Ioannidis2009-hv, bringing into question the reliability of
those studies.
Workshops conducted by NIH intramural researchers in 2018/2019 sought to reproduce five bioinformatics
studies and could not reproduce *any*, citing missing data, software and documentation
@Zaringhalam2020-ms.
A recent systematic analysis of Jupyter notebooks in biomedical articles had similar observations, with
only 245/4169 notebooks (5.9%) giving similar results compared to the original, with irreproducibility
blamed on missing data, broken dependencies and buggy code @Samuel2022-ha.
A similar survey of R scripts in the Harvard Dataverse repository found slightly better results,
with 26% of scripts completing without errors [@Trisovic2022-dq].
Irreproducible bioinformatics has the potential to place patient safety at risk, as demonstrated by an
earlier case study @Baggerly2010-nr.
Researchers are theoretically able to achieve 100% computational reproducibility, yet as we see,
in practice this is rarely achieved.

(TODO: discuss the ramifications of such shortcomings on scientific progress.
This will highlight the urgency and significance of the proposed measures.)

(Unreliable research is a waste of resources and time and can mislead readers to pursue doomed
research directions.)

@@

>>>>

Much has been written on how irreproducibility has reached crisis levels in psychology and life
sciences [@Baker2015; @Begley2012-di].
Indeed, in bioinformatics, the situation is grim with a 2009 systematic evaluation showing only
2 of 18 articles could be reproduced (11%) @Ioannidis2009-hv, bringing into question the reliability of
those studies.
Workshops conducted by NIH intramural researchers in 2018/2019 sought to reproduce five bioinformatics
studies and could not reproduce *any*, citing missing data, software and documentation
@Zaringhalam2020-ms.
A recent systematic analysis of Jupyter notebooks in biomedical articles had similar observations, with
only 245/4169 notebooks (5.9%) giving similar results compared to the original, with irreproducibility
blamed on missing data, broken dependencies and buggy code @Samuel2022-ha.
A similar survey of R scripts in the Harvard Dataverse repository found slightly better results,
with 26% of scripts completing without errors [@Trisovic2022-dq].

The ramifications of unreliable research include misleading the community, wasting research funds,
slowing scientific progress, eroding public confidence in science and tarnishing the reputation
of associated institutions and colleagues.
In certain cases, irreproducible bioinformatics has the potential to place patient safety at risk.

For example in 2006, an article entitled "Genomic signatures to guide the use of chemotherapeutics"
generated a great deal of interest as it was an early application of high throughput transcriptomics
in the prediction of individual patient responses to different chemotherapies @Potti2006-yk.
After identifying some unusual features of the patient group, Baggerly and Coombes attempted
reproduction of some of the key findings @Baggerly2010-nr.
Without the help of scripted workflows to guide the re-analysis, the team used forensic bioinformatic
techniques to piece together how the dataset was originally analysed.
Their investigations found a litany of issues.
Firstly, the labeling of patients in the test set as "responders" or
"non-responders" had been reversed in the original.
Secondly, some of the patients were included more than once (some up to four times) in the analysis,
which is likely to lead to major distortions in results.
Confusingly, some of the reused data has inconsistent grouping; ie: some of the samples are labeled
both sensitive and resistant.
Additional errors include two cases of reported results for the wrong drug.
Baggerly and Coombes highlight that such mistakes can inadvertently occur when using spreadsheets, and
they note that poor documentation and lack of code contributed to a lack of transparency and poor data
hygiene.
The article underwente two corrigenda, but was untimately retracted in 2011, as the authors were not
able to reproduce the findings themselves due to "corruption of several validation data sets"
@Potti2011-ut.
As the array findings were the basis for clinical trials where patients were allocated to treatments,
the flawed data analysis may have had damaging consequences to patients given the wrong drug in the
period 2007-2010, before Duke terminated the trials and suspended the lead author, Dr Anil Potti, who
later resigned.
Duke was served eight lawsuits by families of affected patients seeking compensation for exposure to
harmful and unnecessary chemotherapy, which were settled out of court @Dyer2015-my.

In order for bioinformatic-driven findings to be tranalated to the clinic, they need to be

as demonstrated by an earlier case study @Baggerly2010-nr. <more detail here>

These consequences highlight the urgency for which we must improve stadards for computational research

@@

https://bmcresnotes.biomedcentral.com/articles/10.1186/s13104-021-05875-3#Sec2
